---
title: "Project 1"
author: "Hien Bui, Ana Martinez, Bryanna Partida"
date: "2025-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# loading necessary libraries
library(dplyr)
library(geosphere)
library(kableExtra)
```


```{r echo=FALSE}
# reading the data sets
confirmed_df <- read.csv("time_series_covid19_confirmed_global.csv")
deaths_df <- read.csv("time_series_covid19_deaths_global.csv")

#summary(confirmed_df)
#summary(deaths_df)
```

## **Objective 1:Origin**
*Private Diagnostic Laboratories*
```{r}
#finding columns for the first day
confirmed_first_day <- confirmed_df %>%
  select(`Province.State`, `Country.Region`, X1.22.20)

#then the maximum number of confirmed cases on the first day
max_confirmed <- max(confirmed_first_day$X1.22.20, na.rm = TRUE)

#filter the rows with the maximum confirmed cases
origin_confirmed <- confirmed_first_day %>%
  filter(X1.22.20 == max_confirmed)

print("Area with the most confirmed cases on 2020-01-22:")
print(origin_confirmed)
```
*Global Health Union*
```{r}
#now we do it for deaths
deaths_first_day <- deaths_df %>%
  select(`Province.State`, `Country.Region`, X1.22.20)

max_deaths <- max(deaths_first_day$X1.22.20, na.rm = TRUE)

origin_deaths <- deaths_first_day %>%
  filter(X1.22.20 == max_deaths)

print("Area with the most deaths on 2020-01-22:")
print(origin_deaths)
```
*Origin*
```{r}
#writing prediction of the origin
predicted_origin_confirmed_province<-origin_confirmed$`Province.State`
predicted_origin_confirmed_country <-origin_confirmed$`Country.Region`
predicted_origin_deaths_province <- "Hubei"
predicted_origin_deaths_country <- "China"

if (predicted_origin_confirmed_province == predicted_origin_deaths_province &&
    predicted_origin_confirmed_country == predicted_origin_deaths_country) {
  print(paste("Based on the greatest number of confirmations and deaths on January 22, 2020,",
              predicted_origin_confirmed_province, ",", predicted_origin_confirmed_country,
              "is predicted to be the origin of COVID-19."))
} else {
  print(paste("The area with the most confirmed cases on January 22, 2020, was:",
              predicted_origin_confirmed_province, ",", predicted_origin_confirmed_country))
  print(paste("The area with the most deaths on January 22, 2020, was:",
              predicted_origin_deaths_province, ",", predicted_origin_deaths_country))
  print("Based on these findings, we can infer the potential origin.")
}
```

  ***We used these data sets of Covid-19 of confirmations and deaths to searched for the earliest recorded day to determine the day 1 of the infection. After we found that January 22,2020 was the first day, we proceeded to figure out the number of confirmed cases and deaths on that day. It was Hubei province of China that had the highest number of confirmed cases and death. And with that information, we were able to determine that Hubei was most likely the origin of the Covid-19 according to this data sets.***

## **Objective 2: Most recent area**

```{r}

#making variables to store the most recent first case information
date_columns <- names(confirmed_df)[5:ncol(confirmed_df)]
most_recent_first_case_date <- as.Date("2020-01-21") 
#and initializing date before the first recorded date
most_recent_first_case_province <- NA
most_recent_first_case_country <- NA

#loop through each row of the confirmed_df and the date columns
for (i in 1:nrow(confirmed_df)) {
  province <- confirmed_df$`Province.State`[i]
  country <- confirmed_df$`Country.Region`[i]

  first_case_found <- FALSE
  first_case_date <- NA


  for (date_col in date_columns) {
    cases <- confirmed_df[i, date_col]
    if (!is.na(cases) && cases > 0) {
      first_case_date <- as.Date(gsub("X", "", date_col), format = "%m.%d.%y")
      first_case_found <- TRUE
      break 
#break from the inner loop once the first case is found for this location
    }
  }

#updating the most recent first case if a later date is found
  if (first_case_found && first_case_date > most_recent_first_case_date) {
    most_recent_first_case_date <- first_case_date
    most_recent_first_case_province <- province
    most_recent_first_case_country <- country
  }
}

print("The most recent area to have a first confirmed case:")
print(paste("Province/State:", most_recent_first_case_province))
print(paste("Country/Region:", most_recent_first_case_country))
print(paste("Date:", format(most_recent_first_case_date, "%Y-%m-%d")))
```

***We were able to identify that Pitcairn Islands of the United Kingdom is the most recent area to have a first confirmed case. By making a loop to iterates through the data sets to find the locations where cases changed to 0 to positive. And when combined with the if statements, we  can systematically examine each date's new case reports.***

## **Objective 3: Distance from the origin**

```{r}
# Coordinates of Wuhan, China (Origin)
origin_lon <- 112.2707
origin_lat <- 30.9756
origin_coords <- c(origin_lon, origin_lat)

# Coordinates of Pitcairn Islands, UK (Most Recent First Case)
recent_lon <- -130.1003
recent_lat <- -25.0407
recent_coords <- c(recent_lon, recent_lat)

# Calculate the distance in meters using distm
distance_meters <- geosphere::distm(origin_coords, recent_coords)
print(paste("Distance in meters:", distance_meters))

# Convert meters to miles
distance_miles <- distance_meters * 0.000621371
print(paste("Distance in miles:", distance_miles))

# Format the output string
recent_region <- "Pitcairn Islands, United Kingdom"
origin_location <- "Wuhan, China"
distance_miles_rounded <- round(distance_miles, 2) # Round to 2 decimal places

output_statement <- paste(recent_region, "is", distance_miles_rounded, "miles away from", origin_location)
print(output_statement)
```

***By imputing the coordinates of the origin and the most recent reported case. Using the 'geosphere::distance',we got the distance between the Wuhan and Pitcairn Islands. But the result was in 'meters' so then we had to convert it into miles afterwards.***

## **Objective 4: Risk Score Analysis**
```{r}
# Clean Data
clean_deaths_df <- deaths_df %>%
  filter(!(is.na(Long) | is.na(Lat) | Long == 0 | Lat == 0))
clean_confirmed_df <- confirmed_df %>%
  filter(!(is.na(Long) | is.na(Lat) | Long == 0 | Lat == 0))

# Most recent date column
latest_date <- tail(names(clean_confirmed_df), 1)

# Group data by country and sum latest values
total_confirmed <- clean_confirmed_df %>%
  group_by(`Country.Region`) %>%
  summarize(Confirmed = sum(!!sym(latest_date)), .groups = "drop")
total_deaths <- clean_deaths_df %>%
  group_by(`Country.Region`) %>%
  summarize(Deaths = sum(!!sym(latest_date)), .groups = "drop")

# Merge confirmed/death totals into one data frame
risk_df <- merge(total_confirmed, total_deaths, by = "Country.Region")

# Calculate risk score
risk_df <- risk_df %>%
  mutate(RiskScore = round((100 * Deaths / Confirmed), 2)) %>%
  filter(!is.na(RiskScore) & Confirmed > 0)

# Get global risk score
total_global_confirmed <- sum(risk_df$Confirmed)
total_global_deaths <- sum(risk_df$Deaths)
global_risk_score <- round((100 * total_global_deaths / total_global_confirmed), 2)

# Find countries with highest and lowest risk score
lowest_risk_country <- risk_df %>%
  filter(RiskScore == min(RiskScore)) %>%
  arrange(desc(Confirmed)) %>%
  slice(1)
highest_risk_country <- risk_df %>%
  filter(RiskScore == max(RiskScore)) %>%
  arrange(desc(Confirmed)) %>%
  slice(1)

# Output
lowest_risk_country
highest_risk_country
global_risk_score

```

 ***The area of the world with the lowest risk score is _Singapore_, with a risk score of 0.05%, a very low death rate relative to its high number of confirmed cases. On the other hand, the area with the highest risk score is _Yemen_, with a risk score of 29.07%, showing a much higher fatality rate compared to its case count. The global risk score is approximately 1.12%, this places both countries on opposite ends of the spectrum. Calculating risk scores can be helpful for identifying which regions are managing the disease more effectively or see who is under greater threat. Although it is important to consider that the metric has limitations. This metric does not account for testing availability, healthcare access, population demographics, all of which can greatly affect how the data is interpreted***

## **Objective 5: Top 5 Countries by Confirmations and Deaths**
```{r}
# Most recent date column in the dataset
latest_date <- tail(names(confirmed_df), 1)

# Group confirmed cases by Country/Region and sum
confirmed_totals <- confirmed_df %>%
  group_by(Country.Region) %>%
  summarise(Total_Confirmed = sum(!!sym(latest_date))) %>%
  arrange(desc(Total_Confirmed)) %>%
  head(5)

# Group deaths by Country/Region and sum
deaths_totals <- deaths_df %>%
  group_by(Country.Region) %>%
  summarise(Total_Deaths = sum(!!sym(latest_date))) %>%
  arrange(desc(Total_Deaths)) %>%
  head(5)

# The top 5 countries by confirmations
print("Top 5 Countries by Confirmed Cases")
confirmed_totals %>%
  kable(caption = paste("Top 5 Countries with Most Confirmed COVID-19 Cases as of", latest_date)) %>%
  kable_styling(full_width = FALSE)

# The top 5 countries by deaths
print("Top 5 Countries by Deaths")
deaths_totals %>%
  kable(caption = paste("Top 5 Countries with Most COVID-19 Related Deaths as of", latest_date)) %>%
  kable_styling(full_width = FALSE)

```


